---
title: "Foreclosure Data Analysis"
author: "Jasper Evans, Krehl Kasayan, Paul Kiefer, Pablo Suarez"
date: "2024-20-2024"
output: html_document
---

# Prince George's County Maryland Foreclosure Analysis

We will use Prince George's County's foreclosure dataset as the core of our project (https://data.princegeorgescountymd.gov/Urban-Planning/County-Foreclosures/mnie-hrv7/about_data). The dataset is maintained by the Prince George's County planning department and lists foreclosures by address between 2009 and 2024. 

## Data Setup

Load libraries.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
#install.packages("ggrepel")
#install.packages('ggthemes')
library(tidyverse)
library(lubridate)
library(janitor)
library(ggthemes)
library(tidycensus)
library(ggplot2)
library(ggrepel)
library(tigris)
library(sf)
```

The foreclosure dataset contains 71,676 rows and 10 columns, including the street number, street name, city, state and zip code of each foreclosure.

The data was then cleaned and a column was created to concatenate the street number, street name, city, state and zip code information into a new column called full_address.
```{r}
foreclosure_pg <- read_csv("datasets/pg_foreclosures.csv", guess_max = 71677)
```