---
title: "Foreclosure Data Analysis"
author: "Jasper Evans, Krehl Kasayan, Paul Kiefer, Pablo Suarez"
date: "2024-20-2024"
output: html_document
---

# Prince George's County Maryland Foreclosure Analysis

We will use Prince George's County's foreclosure dataset as the core of our project (https://data.princegeorgescountymd.gov/Urban-Planning/County-Foreclosures/mnie-hrv7/about_data). The dataset is maintained by the Prince George's County planning department and lists foreclosures by address between 2009 and 2024. 

## Data Setup

Load libraries.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
##install.packages("ggrepel")
##install.packages('ggthemes')
library(tidyverse)
library(lubridate)
library(janitor)
library(ggthemes)
library(tidycensus)
library(ggplot2)
library(ggrepel)
library(tigris)
library(sf)
library(ggmap)
```


The foreclosure dataset contains 71676 rows and 10 columns, including the street number, street name, city, state and zip code of each foreclosure.

```{r}
foreclosure_pg <- read_csv("datasets/pg_foreclosures.csv", guess_max = 71677)
```

We will eventually need to generate census tract ID numbers for each address, so we will create a new dataframe that excludes records with missing zip code values. This filters out roughly 1600 records.

```{r}
foreclosure_pg_census_tracts <- read_csv("datasets/pg_foreclosure_tract_geocodio.csv", guess_max = 71677) |>
  clean_names()
```

```{r}
tracts_unique <- length(unique(foreclosure_pg_census_tracts$census_tract_code))

print(tracts_unique)
```

```{r}
# Perform inner join based on the 'location' column
pg_foreclosures_with_tracts <- left_join(foreclosure_pg_census_tracts, foreclosure_pg, by = "location")

# Print the merged dataset
print(pg_foreclosures_with_tracts)
```

```{r}
pg_foreclosures_filtered <- pg_foreclosures_with_tracts[!is.na(pg_foreclosures_with_tracts$census_tract_code), ]
```

```{r}
pg_foreclosures_filtered <- select(pg_foreclosures_filtered, -accuracy_score, -accuracy_type, -source, -full_fips_block, -metro_micro_statistical_area_name,-metro_micro_statistical_area_code, -metro_micro_statistical_area_type, -combined_statistical_area_name, -metropolitan_division_area_name, -metropolitan_division_area_code, -addressoccupied, -city.y, -state.y, street_address)
```

